# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
Data contains many "NA" values, for some days all data points are "NA"s  
the intervals are five minute time intervals spaning an entire day.  They are coded "hhmm", so for example  1020 is 10:20am and 1835 is 18:35 in a 24 hour clock or 6:35 pm
```{r}
library(knitr)
rawData <- data.frame(read.csv("activity.csv"))
#data frame with no NAs, this will leave some days with no values
DataNoNAs <-rawData[complete.cases(rawData),]
```


## What is mean total number of steps taken per day?
The assignment states the "NA" values can be ignored for this calcualtion, and this is what was done.  Days with "NA"'s for all data points for that day are reported as 0 steps per day, and teh average and median as 0 also.


```{r} 

#split(rawData,list(rawData$date))
sumPerDay <- sapply(split(rawData,list(rawData$date)),function(x) sum(x$steps,na.rm=TRUE) )
hist(sumPerDay,breaks=30,xlab="Total Steps Per day",main="Histogram: Number of Steps per Day, 2012-10-01 through 2012-11-30" )
meanPerDay <- mean(sumPerDay)
medianPerDay <- median(sumPerDay)

```
### The mean total number of steps per day is `r meanPerDay `
### The median total number of steps per day is `r medianPerDay `

## What is the average daily activity pattern?



```{r}
#split data by interval and calculate means, remove NAs
meanPerInterval <- sapply(split(rawData,list(rawData$interval)),function(x) mean(x$steps,na.rm=TRUE) )
#extract column names for graph
xLabels <- names(meanPerInterval)

#Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

#plot data
plot(meanPerInterval,type="l",main="Mean Number of Steps per Interval",xlab="Interval",ylab="Mean Number of Steps", xaxt="n")
#do calculations for x axis
xLabIndx = seq(1,288,by=10)
#add x axis labels 
axis(at=xLabIndx,side=1,labels=xLabels[xLabIndx],las=2)

#Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?


max(meanPerInterval)
#max doesn't return which interval, so using sort
sort(meanPerInterval,decreasing=TRUE)[1]

```

### The 5-minute interval that on average has the maximum number of steps is `r names(sort(meanPerInterval,decreasing=TRUE)[1]) ` and the number of average steps for this interval is `r sort(meanPerInterval,decreasing=TRUE)[1] ` 



## Imputing missing values

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}

#check to see whre the NAs are
sum(is.na(rawData$steps))
sum(is.na(rawData$date))
sum(is.na(rawData$interval))
#they are only in the setps data


```


Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

Strategy:  will replace NA values with median value for that interval accross all days. 

Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}

#Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?


#create vector of median values for intervals
medianPerInterval <- sapply(split(DataNoNAs,list(DataNoNAs$interval)),function(x) median(x$steps,na.rm=TRUE) )
#replace the "NA"'s, first copy rawData
replaceNAs.df <- rawData
#mkake index of NAs
indx <- which(is.na(replaceNAs.df$steps==TRUE))
#using index, replase NAs with median values
for (i in 1:length(indx) ) {
replaceNAs.df[indx[i],1]  = medianPerInterval[[as.character(replaceNAs.df[indx[i],3])]]
}
sumPerDay.ReplaceNAs <- sapply(split(replaceNAs.df,list(replaceNAs.df$date)),function(x) sum(x$steps,na.rm=TRUE) )
hist(sumPerDay.ReplaceNAs,breaks=30,xlab="Total Steps Per day",main="Histogram: Number of Steps per Day, 2012-10-01 through 2012-11-30" )
meanPerDay.ReplaceNAs <- mean(sumPerDay.ReplaceNAs)
medianPerDay.ReplaceNAs <- median(sumPerDay.ReplaceNAs)
```

- The mean total number of steps per day after replacing NAs is `r meanPerDay.ReplaceNAs ` , without repalcing NAs the mean is `r meanPerDay `  
- The median total number of steps per day after replacing NAs is `r medianPerDay.ReplaceNAs ` , without replacing NAs the medain is `r medianPerDay `.  


### Imputing values for NAs by this method has an effect on the mean but not the median for this data set





## Are there differences in activity patterns between weekdays and weekends?
```{r, fig.height=5,fig.width=10}

#For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

#Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
#Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was creating using simulated data:
library(ggplot2)
#add weekend/weekday factor
isWeekend <- weekdays(as.Date(replaceNAs.df$date)) %in% c("Sunday","Saturday")
replaceNAs.df$weekend[isWeekend] <- "weekend"
replaceNAs.df$weekend[!isWeekend] <- "weekday"
#plot
g <-ggplot(replaceNAs.df,aes(interval,steps))
g  + facet_grid(weekend ~ .) + 
    stat_summary(fun.y = "mean", colour = "red",geom="line") +
    labs(title="Activty Patterns for Weekday vs. Weekend") + labs(y="Mean number of steps")





```

### The data show there is difference between weekday and weekend activty. On average, the is early activity on weekdays, while weekends have a higher level of activity in mid-day and late day intervals.